GAM exploration

### 2024-05-03 manu-draft.qmd
light attenuation added to fig 2, caption updated

![Seagrass changes over time in Tampa Bay for (a) areal coverage 1988 - 2022 from mapping and (b) frequency occurrence of major species 1998 - 2022 from annual transect monitoring. Changes are shown for major bay segments. Red lines in (a) show approximate capacity of seagrass coverage based on the baywide target of 40,000 acres. OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/seagrasschg.png){#fig-seagrasschg}

>>>>

![Seagrass changes over time in Tampa Bay for (a) areal coverage 1988 - 2022 from mapping, (b) frequency occurrence of major species 1998 - 2022 from annual transect monitoring, and (c) mean annual light attenuation from 1970 to 2022. Changes are shown for major bay segments. Red lines in (a) show approximate capacity of seagrass coverage based on the baywide target of 40,000 acres and red lines in (c) show the approximate linear trend for the period of record covered by the transect data in (b). The dashed line in (c) shows the light attenuation threshold for each bay segment identified as supportive for seagrass growth. OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/seagrasschg.png){#fig-seagrasschg}

### 2024-05-03 revision/gam_exploration.qmd
light attenuation added to fig 2, caption updated

The following is an exploratory analysis of potential seagrass response related to time, light attenuation, temperature, and salinity. Three different datasets are used, each with different sampling designs. GAMs are used to assess potential non-linear responses in seagrass change using the sum of additive smoothing functions for different predictors. Lower Tampa Bay is not modeled in any of the models given minimal seagrass loss.  Model selection techniques are not considered for simplicity and the results are evaluated based on the percent deviance explained and "partial effects" of each smoother, where the smoother is a proxy for the effect of the variable on the response while considering a more comprehensive set of predictors.  These models were developed in response to reviewer comments on including additional predictors important for characterizing change, namely time and light attenuation.

@@

* Seagrass and stressor data are collected at the same locations possibly allowing for a more consistent comparison
* The stressor data (light attenuation, temperature, salinity) are observed data, so the expectation of how seagrass will response varies from the EPC data and by stressor, i.e., seagrass will decrease with temperature and light attenuation, but increase with salinity (fresh conditions are more stressful)
* Monthly observations are available, allowing a seasonal component in the model. 
* Secchi data is used for light attenuation (as for EPC), but more than half of the observations are on the bottom.  Attempts were made to calculate monthly medians using Kaplan-Meier survival curves, but data were insufficient to do so.

@@

```{r}
#| fig-width: 8
#| fig-height: 3
draw(mod, select = 's(temp)', ncol = 3, partial_match = TRUE)
draw(mod, select = 's(sal)', ncol = 3, partial_match = TRUE)
```
```{r}
#| fig-width: 10
#| fig-height: 4
draw(mod, select = 'ti(temp,mo)', ncol = 3, partial_match = TRUE)
draw(mod, select = 'ti(sal,mo)', ncol = 3, partial_match = TRUE)
```

@@

```{r}
#| fig-width: 7
#| fig-height: 3
tempslc <- smooth_estimates(mod, select = 'ti(temp,mo)', partial_match = T) %>% 
  add_confint() %>% 
  filter(mo %in% c(1, 7)) %>% 
  filter(bay_segment %in% c('HB', 'MTB'))
ggplot(tempslc, aes(x = temp, y = .estimate, color = factor(mo), fill = factor(mo), group = mo)) +
  geom_ribbon(aes(ymin = .lower_ci, ymax = .upper_ci), alpha = 0.2, color = NA) +
  geom_line() +
  facet_wrap(~bay_segment, scales = 'free_y') + 
  labs(
    x = 'Temperature',
    y = 'Partial effect',
    color = 'Month',
    fill = 'Month'
  )
```
```{r}
#| fig-width: 4
#| fig-height: 3
salslc <- smooth_estimates(mod, select = 'ti(sal,mo)', partial_match = T) %>% 
  add_confint() %>% 
  filter(mo %in% c(3, 9)) %>% 
  filter(bay_segment %in% c('MTB'))
ggplot(salslc, aes(x = sal, y = .estimate, color = factor(mo), fill = factor(mo), group = mo)) +
  geom_ribbon(aes(ymin = .lower_ci, ymax = .upper_ci), alpha = 0.2, color = NA) +
  geom_line() +
  facet_wrap(~bay_segment, scales = 'free_y') + 
  labs(
    x = 'Salinity',
    y = 'Partial effect',
    color = 'Month',
    fill = 'Month'
  )
```

@@

$$
\begin{aligned}
Freq. Occ. \sim s(year, by = stratum, bs = 'tp') + s(la, by = stratum, bs = 'tp') + 
\\
s(temp, by = stratum, bs = 'tp') + s(salinity, by = stratum, bs = 'tp') +
\\
s(month, bs = 'cc', by = stratum) + 
\\
ti(year, month, bs = c('tp', 'cc'), by = stratum) + 
\\
ti(la, month, bs = c('tp', 'cc'), by = stratum) + 
\\
ti(temp, month, bs = c('tp', 'cc'), by = stratum) +
\\
ti(salinity, month, bs = c('tp', 'cc'), by = stratum)
\end{aligned}
$$
```{r}

@@

```{r}
#| fig-width: 8
#| fig-height: 3
draw(mod, select = 's(temp)', ncol = 4, partial_match = TRUE)
draw(mod, select = 's(sal)', ncol = 4, partial_match = TRUE)
```
```{r}
#| fig-width: 10
#| fig-height: 4
draw(mod, select = 'ti(temp,mo)', ncol = 4, partial_match = TRUE)
draw(mod, select = 'ti(sal,mo)', ncol = 4, partial_match = TRUE)
```

@@

```{r}
#| fig-width: 7
#| fig-height: 3
tempslc <- smooth_estimates(mod, select = 'ti(temp,mo)', partial_match = T) %>% 
  add_confint() %>% 
  filter(mo %in% c(1, 7)) %>% 
  filter(site %in% c('E1', 'E4'))
ggplot(tempslc, aes(x = temp, y = .estimate, color = factor(mo), fill = factor(mo), group = mo)) +
  geom_ribbon(aes(ymin = .lower_ci, ymax = .upper_ci), alpha = 0.2, color = NA) +
  geom_line() +
  facet_wrap(~site, scales = 'free_y') + 
  labs(
    x = 'Temperature',
    y = 'Partial effect',
    color = 'Month',
    fill = 'Month'
  )
```
```{r}
#| fig-width: 4
#| fig-height: 3
salslc <- smooth_estimates(mod, select = 'ti(sal,mo)', partial_match = T) %>% 
  add_confint() %>% 
  filter(mo %in% c(3, 9)) %>% 
  filter(site %in% c('E4'))
ggplot(salslc, aes(x = sal, y = .estimate, color = factor(mo), fill = factor(mo), group = mo)) +
  geom_ribbon(aes(ymin = .lower_ci, ymax = .upper_ci), alpha = 0.2, color = NA) +
  geom_line() +
  facet_wrap(~site, scales = 'free_y') + 
  labs(
    x = 'Salinity',
    y = 'Partial effect',
    color = 'Month',
    fill = 'Month'
  )
```

>>>>

The following is an exploratory analysis of potential seagrass response related to time, light attenuation, temperature, and salinity. Three different datasets are used, each with different sampling designs. GAMs are used to assess potential non-linear responses in seagrass change using the sum of additive smoothing functions for different predictors. Lower Tampa Bay is not modeled in any of the models given minimal seagrass loss. Model selection techniques are not considered for simplicity and the results are evaluated based on the percent deviance explained and "partial effects" of each smoother, where the smoother is a proxy for the effect of the variable on the response while considering a more comprehensive set of predictors.

@@

-   Seagrass and stressor data are collected at the same locations possibly allowing for a more consistent comparison
-   The stressor data (light attenuation, temperature, salinity) are observed data, so the expectation of how seagrass will response varies from the EPC data and by stressor, i.e., seagrass will decrease with temperature and light attenuation, but increase with salinity (fresh conditions are more stressful)
-   Monthly observations are available, allowing a seasonal component in the model.
-   Secchi data is used for light attenuation (as for EPC), but more than half of the observations are on the bottom. Attempts were made to calculate monthly medians using Kaplan-Meier survival curves, but data were insufficient to do so.

@@

```{r}
#| fig-width: 8
#| fig-height: 3
draw(mod, select = 's(temp)', ncol = 3, partial_match = TRUE)
draw(mod, select = 's(sal)', ncol = 3, partial_match = TRUE)
```

```{r}
#| fig-width: 10
#| fig-height: 4
draw(mod, select = 'ti(temp,mo)', ncol = 3, partial_match = TRUE)
draw(mod, select = 'ti(sal,mo)', ncol = 3, partial_match = TRUE)
```

@@

```{r}
#| fig-width: 7
#| fig-height: 3
tempslc <- smooth_estimates(mod, select = 'ti(temp,mo)', partial_match = T) %>% 
  add_confint() %>% 
  filter(mo %in% c(1, 7)) %>% 
  filter(bay_segment %in% c('HB', 'MTB'))
ggplot(tempslc, aes(x = temp, y = .estimate, color = factor(mo), fill = factor(mo), group = mo)) +
  geom_ribbon(aes(ymin = .lower_ci, ymax = .upper_ci), alpha = 0.2, color = NA) +
  geom_line() +
  facet_wrap(~bay_segment, scales = 'free_y') + 
  labs(
    x = 'Temperature',
    y = 'Partial effect',
    color = 'Month',
    fill = 'Month'
  )
```

```{r}
#| fig-width: 4
#| fig-height: 3
salslc <- smooth_estimates(mod, select = 'ti(sal,mo)', partial_match = T) %>% 
  add_confint() %>% 
  filter(mo %in% c(3, 9)) %>% 
  filter(bay_segment %in% c('MTB'))
ggplot(salslc, aes(x = sal, y = .estimate, color = factor(mo), fill = factor(mo), group = mo)) +
  geom_ribbon(aes(ymin = .lower_ci, ymax = .upper_ci), alpha = 0.2, color = NA) +
  geom_line() +
  facet_wrap(~bay_segment, scales = 'free_y') + 
  labs(
    x = 'Salinity',
    y = 'Partial effect',
    color = 'Month',
    fill = 'Month'
  )
```

@@

$$
\begin{aligned}
Freq. Occ. \sim s(year, by = stratum, bs = 'tp') + s(la, by = stratum, bs = 'tp') + 
\\
s(temp, by = stratum, bs = 'tp') + s(salinity, by = stratum, bs = 'tp') +
\\
s(month, bs = 'cc', by = stratum) + 
\\
ti(year, month, bs = c('tp', 'cc'), by = stratum) + 
\\
ti(la, month, bs = c('tp', 'cc'), by = stratum) + 
\\
ti(temp, month, bs = c('tp', 'cc'), by = stratum) +
\\
ti(salinity, month, bs = c('tp', 'cc'), by = stratum)
\end{aligned}
$$

```{r}

@@

```{r}
#| fig-width: 8
#| fig-height: 3
draw(mod, select = 's(temp)', ncol = 4, partial_match = TRUE)
draw(mod, select = 's(sal)', ncol = 4, partial_match = TRUE)
```

```{r}
#| fig-width: 10
#| fig-height: 4
draw(mod, select = 'ti(temp,mo)', ncol = 4, partial_match = TRUE)
draw(mod, select = 'ti(sal,mo)', ncol = 4, partial_match = TRUE)
```

@@

```{r}
#| fig-width: 7
#| fig-height: 3
tempslc <- smooth_estimates(mod, select = 'ti(temp,mo)', partial_match = T) %>% 
  add_confint() %>% 
  filter(mo %in% c(1, 7)) %>% 
  filter(site %in% c('E1', 'E4'))
ggplot(tempslc, aes(x = temp, y = .estimate, color = factor(mo), fill = factor(mo), group = mo)) +
  geom_ribbon(aes(ymin = .lower_ci, ymax = .upper_ci), alpha = 0.2, color = NA) +
  geom_line() +
  facet_wrap(~site, scales = 'free_y') + 
  labs(
    x = 'Temperature',
    y = 'Partial effect',
    color = 'Month',
    fill = 'Month'
  )
```

```{r}
#| fig-width: 4
#| fig-height: 3
salslc <- smooth_estimates(mod, select = 'ti(sal,mo)', partial_match = T) %>% 
  add_confint() %>% 
  filter(mo %in% c(3, 9)) %>% 
  filter(site %in% c('E4'))
ggplot(salslc, aes(x = sal, y = .estimate, color = factor(mo), fill = factor(mo), group = mo)) +
  geom_ribbon(aes(ymin = .lower_ci, ymax = .upper_ci), alpha = 0.2, color = NA) +
  geom_line() +
  facet_wrap(~site, scales = 'free_y') + 
  labs(
    x = 'Salinity',
    y = 'Partial effect',
    color = 'Month',
    fill = 'Month'
  )
```
